---
title: "Data Analysis and the English 2019/20 Premier League Season"
author: "Leon Benito"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: spacelab
    highlight-style: tango
    embed-resources: true
    fig-width: 8
    fig-height: 8
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

invisible(lapply(c("tidyverse", "here", "readxl", "DT", "ggimage", 
                   "ggcorrplot", "ggrepel", "patchwork", "knitr",
                   "gt", "gtExtras", "showtext"), 
                 library, character.only = TRUE))

font_add_google("Roboto Condensed", "roboto")
showtext_auto()

```

# Introduction

No statistical analysis can do justice to the excitement of the 2019/20 Premier League season. Without any coincidence, the season was won by Liverpool. Here, I analyse match data from every teams' matches, capturing results, aspects of performance including both attack and defense, as well as discipline. Liverpool were unsurpassed in almost every dimension. On 99 points for the season, Liverpool, beat Manchester City by 18 points and the third and fourth placed teams by 33 points.

# The raw data

After importing the raw data, the first task is to present some basic summary statistics on the structure of the dataframe and its contents.

The table below 'glimpses' the dataframe itself. I then summarise some basic summary statistics.

```{r}
df <- read_excel(here("data", "premierLeague.xlsx")) 

glimpse(df)
print(paste("No of observations: ", nrow(df)))
print(paste("No of variables: ", ncol(df)))

```

The `r nrow(df)` observations in the dataframerefer to match-level data for the 2019/20 season: each of 20 teams playing 38 matches in the season. For each match, we have a total of `r ncol(df)` variables.

# Some simple correlations

A useful next stage is to inspect correlations among the 24 numeric variables. These correlations are shown below.

Not surprisingly, 'possession' indicators (possession, touches, passes, successful passes, corners, crosses) are positively correlated. These tend to be are negatively correlated with saves, clearances and goal kicks (defensive acts).

```{r}
# Calculate correlation matrix
cor_matrix <- df %>% 
  select_if(is.numeric) %>% 
  select(-Penalties) %>%
  cor(use = "complete.obs")

# Create improved correlation plot
ggcorrplot(cor_matrix,
           method = "square",
           type = "upper",
           lab = FALSE,
           colors = c("#8B4513", "#FFFAF0", "#006400"),  # Stronger contrast
           outline.color = "gray60",
           ggtheme = theme_minimal(base_family = "roboto"),
           legend.title = "Correlation") +
  theme(
    text = element_text(family = "roboto"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, 
                               size = 12, family = "roboto", face = "bold"),
    axis.text.y = element_text(angle = 0, hjust = 1, 
                               size = 12, family = "roboto", face = "bold"),
    legend.text = element_text(family = "roboto", size = 11),
    legend.title = element_text(family = "roboto", size = 12, face = "bold"),
    panel.grid.major = element_blank(),
    plot.margin = margin(10, 10, 10, 20)
  )
```

# Performance

## The league table

The measure of performance that ultimately matters is the position of the table at the end of the season. That said, the excitement is in the journey of getting there.

```{r}
league_table <- df %>%
  group_by(Match_Name) %>%
  mutate(
    OpponentScore = ifelse(row_number() == 1, Score[2], Score[1]),
    Result = case_when(
      Score > OpponentScore ~ "W",
      Score < OpponentScore ~ "L",
      Score == OpponentScore ~ "D"
    )
  ) %>%
  ungroup() %>%
  group_by(Team) %>%
  summarise(
    Played = n(),
    Won = sum(Result == "W"),
    Drawn = sum(Result == "D"),
    Lost = sum(Result == "L"),
    GF = sum(Score),
    GA = sum(OpponentScore)
  ) %>%
  mutate(
    GD = GF - GA,
    Points = (Won * 3) + Drawn
  ) %>%
  arrange(desc(Points), desc(GD), desc(GF)) %>%
  mutate(Position = row_number()) %>%
  select(Position, Team, Played, Won, Drawn, Lost, GF, GA, GD, Points)

# Display as formatted table
league_table %>%
  gt() %>%
  tab_header(
    title = md("**Premier League 2019/20 Final Table**")
  ) %>%
  opt_table_font(font = google_font("Roboto Condensed")) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 14,
    data_row.padding = px(3),
    column_labels.padding = px(4),
    table.width = pct(85)
  ) %>%
  cols_align(
    align = "center",
    columns = c(Position, Played, Won, Drawn, Lost, GF, GA, GD, Points)
  ) %>%
  cols_align(
    align = "left",
    columns = Team
  ) %>%
  tab_style(
    style = cell_fill(color = "#C8E6C9"),
    locations = cells_body(rows = Position <= 4)
  ) %>%
  tab_style(
    style = cell_fill(color = "#E8F5E9"),
    locations = cells_body(rows = Position == 5)
  ) %>%
  tab_style(
    style = cell_fill(color = "#FFCDD2"),
    locations = cells_body(rows = Position >= 18)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = c(Position, Team, Points))
  )

```

### Attacking statistics

```{r}
df %>%
  select(Team, contains("Shot")) %>%
  group_by(Team) %>%
  summarise(across(everything(), ~mean(., na.rm = TRUE))) %>%
  arrange(desc(ShotsAccuracy)) %>%
  gt() %>%
  tab_header(
    title = "Attacking Statistics, by Team",
    subtitle = "Premier League 2019/20"
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 3
  ) %>%
  opt_table_font(font = google_font("Roboto Condensed")) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 11,
    data_row.padding = px(3),
    column_labels.padding = px(4),
    table.width = pct(80)
  ) %>%
  cols_align(
    align = "center",
    columns = where(is.numeric)
  )


```

## Shot and save accuracy

Shot and save accuracy are standard measures in football sports analytics as an intermediate measure of how team performance.

```{r}

# Create data
team_accuracy <- df %>%
  select(Team, ShotsAccuracy, SavesAccuracy) %>%
  group_by(Team) %>%
  summarise(
    ShotsAccuracy = round(mean(ShotsAccuracy, na.rm = TRUE), 3),
    SavesAccuracy = round(mean(SavesAccuracy, na.rm = TRUE), 3)
  ) %>%
  arrange(desc(ShotsAccuracy))

# Display as formatted table
team_accuracy %>%
  gt() %>%
  tab_header(
    title = md("**Shot and Save Accuracy by Team**"),
    subtitle = "Premier League 2019/20"
  ) %>%
  opt_table_font(font = google_font("Roboto Condensed")) %>%
  tab_options(
    table.font.size = 10,
    heading.title.font.size = 13,
    heading.subtitle.font.size = 10,
    data_row.padding = px(2),
    column_labels.padding = px(3),
    table.width = pct(70),
    heading.padding = px(4)
  ) %>%
  cols_width(
    Team ~ px(150),
    ShotsAccuracy ~ px(120),
    SavesAccuracy ~ px(120)
  ) %>%
  cols_align(
    align = "center",
    columns = c(ShotsAccuracy, SavesAccuracy)
  ) %>%
  cols_align(
    align = "left",
    columns = Team
  ) %>%
  cols_label(
    ShotsAccuracy = "Shot Accuracy",
    SavesAccuracy = "Save Accuracy"
  ) %>%
  fmt_number(
    columns = c(ShotsAccuracy, SavesAccuracy),
    decimals = 3
  )

# Plot 
ggplot(team_accuracy, aes(x = ShotsAccuracy, y = SavesAccuracy)) +
  geom_hline(yintercept = mean(team_accuracy$SavesAccuracy), 
             linetype = "dashed", color = "gray60", alpha = 0.7) +
  geom_vline(xintercept = mean(team_accuracy$ShotsAccuracy), 
             linetype = "dashed", color = "gray60", alpha = 0.7) +
  geom_point(size = 4, color = "#2D5F4F", alpha = 0.8) +
  geom_label_repel(aes(label = Team), 
                   size = 3.5,
                   box.padding = 0.5,
                   point.padding = 0.3,
                   segment.color = "grey50",
                   segment.size = 0.3,
                   fill = "white",
                   alpha = 0.85,
                   max.overlaps = Inf,
                   family = "roboto") +
  scale_x_continuous(expand = expansion(mult = 0.1),
                     breaks = scales::pretty_breaks(n = 8)) +
  scale_y_continuous(expand = expansion(mult = 0.1),
                     breaks = scales::pretty_breaks(n = 8)) +
  labs(title = "Shot Accuracy vs Save Accuracy by Team",
       subtitle = "Premier League 2019/20 | Dashed lines show league averages",
       x = "Shot Accuracy",
       y = "Save Accuracy") +
  theme(base_family = "roboto") +
  theme(
    text = element_text(family = "roboto"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15, family = "roboto"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, family = "roboto", color = "gray40"),
    axis.title = element_text(face = "bold", size = 12, family = "roboto"),
    axis.text = element_text(size = 11, family = "roboto"),
    axis.line = element_line(color = "gray30", linewidth = 0.5),
    axis.ticks = element_line(color = "gray30", linewidth = 0.3),
    axis.ticks.length = unit(0.15, "cm"),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_line(color = "gray95", linewidth = 0.2),
    panel.border = element_rect(color = "gray50", fill = NA, linewidth = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(15, 15, 15, 15)
  )

```

The following density plots capture, team-by-team, shot and save accuracy and how this varies. My impression is this does not vary so much across teams, including between the best and weakest teams. What matters more is the number of shot attempts and preventing an opposing team from shooting.

```{r}

# Prepare data
Acc <- df %>%
  select(Team, ShotsAccuracy, SavesAccuracy) %>%
  arrange(Team)

teams <- unique(Acc$Team)

plots <- lapply(seq_along(teams), function(i) {
  team <- teams[i]
  data <- Acc %>% filter(Team == team)
  
  shot_mean <- mean(data$ShotsAccuracy, na.rm = TRUE)
  save_mean <- mean(data$SavesAccuracy, na.rm = TRUE)
  
  plot_data <- data %>%
    pivot_longer(cols = c(ShotsAccuracy, SavesAccuracy),
                 names_to = "Metric",
                 values_to = "Accuracy")
  
  p <- ggplot(plot_data, aes(x = Accuracy, color = Metric)) +
    geom_density(alpha = 0, linewidth = 1.2) +
    geom_vline(xintercept = shot_mean, color = "#1976D2", alpha = 0.6, linewidth = 1) +
    geom_vline(xintercept = save_mean, color = "#D32F2F", alpha = 0.6, linewidth = 1) +
    annotate("text", x = shot_mean - 0.05, y = 3.8, 
             label = round(shot_mean, 2), color = "#1976D2", 
             vjust = 1, hjust = 1, size = 3, fontface = "bold", family = "roboto") +
    annotate("text", x = save_mean + 0.05, y = 3.8, 
             label = round(save_mean, 2), color = "#D32F2F", 
             vjust = 1, hjust = 0, size = 3, fontface = "bold", family = "roboto") +
    scale_color_manual(values = c("ShotsAccuracy" = "#1976D2", "SavesAccuracy" = "#D32F2F")) +
    scale_x_continuous(limits = c(-0.5, 1.5), breaks = seq(0, 1.5, 0.5)) +
    scale_y_continuous(limits = c(0, 4), breaks = seq(0, 4, 2)) +
    labs(title = team) +
    theme(base_family = "roboto") +
    theme(
      text = element_text(family = "roboto"),
      legend.position = "none",
      axis.title = element_blank(),
      axis.text.x = element_text(size = 8, family = "roboto"),
      axis.text.y = if (i %% 5 == 1) element_text(size = 8, family = "roboto") else element_blank(),
      axis.ticks = element_line(color = "gray40", linewidth = 0.3),
      panel.grid.major.y = element_line(color = "gray90", linewidth = 0.2),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold", family = "roboto"),
      axis.line = element_line(color = "gray40", linewidth = 0.4),
      plot.margin = margin(4, 4, 4, 4)
    )
  
  return(p)
})

combined_plot <- wrap_plots(plots, ncol = 5, nrow = 4) +
  plot_annotation(
    title = "Shot / Save Accuracy",
    subtitle = "Blue = Shots | Red = Saves | Vertical lines show means",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 17, face = "bold", family = "roboto"),
      plot.subtitle = element_text(hjust = 0.5, size = 11, family = "roboto", color = "gray40")
    )
  )
combined_plot

```

# Discipline

Discipline is another feature of performance. On these measures, there was nothing 'cynical' about Liverpool's performances or results. LFC's discipline and sportsmanship was high according to these standard measures.

```{r}
discipline <- df %>% 
  select(Team, Fouls, YellowCard, RedCard, YellowRed) %>% 
  group_by(Team) %>% 
  summarise_at(vars(Fouls, YellowCard:YellowRed), funs(sum)) %>% 
  arrange(Fouls) %>% 
  mutate(TotCards = RedCard+YellowRed) 

disp2 <- discipline %>%  
  arrange(TotCards) %>% 
  filter(Team %in% c("Chelsea", "Manchester United", "Burnley", "Liverpool", "Arsenal", "Manchester City"))

ggplot(discipline, aes(x = Fouls, y = YellowCard, color = TotCards, size = TotCards)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(data = disp2, 
                  mapping = aes(x = Fouls, y = YellowCard, label = Team), 
                  size = 3.5,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = "grey50",
                  segment.size = 0.3,
                  max.overlaps = Inf,
                  inherit.aes = FALSE) +
  scale_color_gradient(low = "#E8DDB5", high = "#8B0000") +
  scale_size_continuous(range = c(2, 8)) +
  guides(size = "none") +
  labs(title = "Premier League Discipline Analysis",
       subtitle = "Relationship between Fouls and Yellow Cards",
       x = "Number of Fouls",
       y = "Yellow Cards",
       color = "Total Cards against",
       caption = "Point size represents number of cards received") +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "grey40"),
    plot.caption = element_text(size = 9, color = "grey50", hjust = 1),
    axis.title = element_text(face = "bold", size = 11),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.3)
  )
```

# Manager of the Season

The upshot of managerial success is to make a team more than the some of its parts. One possibility would be to capture team 'productivity' as a positive deviation from tyhe relation of points relative to wage bill (as a measure of 'inputs'). Our data do not include that wage bill (or total transfer spending) measure of inputs. Past evidence appears to suggest that managers do not add much value relative to wage bill (Kuper and Szymanski, *Soccernomics*).

Without those measure of inputs, and limited by the available data, it is still possible to arrive at indicators of Jurgen Klopp's managerial prowess, including as an attacking force.

```{r}
df %>% 
  select(Team_Logo, Team, Score, Offsides) %>% 
  group_by(Team_Logo, Team) %>% 
  summarise(
    Score = sum(Score, na.rm = TRUE),
    Offsides = sum(Offsides, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = Offsides, y = Score)) +
  geom_image(aes(image = Team_Logo), size = 0.065) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  labs(
    title = "Goals Scored vs Offsides",
    subtitle = "Premier League 2019/20 Season",
    x = "Total Offsides",
    y = "Total Goals Scored"
  ) +
  theme(base_family = "roboto") +
  theme(
    text = element_text(family = "roboto"),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "roboto"),
    plot.subtitle = element_text(size = 11, hjust = 0.5, family = "roboto", color = "gray40"),
    axis.title = element_text(size = 12, face = "bold", family = "roboto"),
    axis.text = element_text(size = 11, family = "roboto"),
    axis.line = element_line(color = "gray40", linewidth = 0.5),
    axis.ticks = element_line(color = "gray40", linewidth = 0.3),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray50", fill = NA, linewidth = 0.5),
    plot.margin = margin(15, 15, 15, 15)
  )
```

```{r}
# Calculate season averages
season_avg_score <- df %>% 
  group_by(Team) %>% 
  summarise(Score = sum(Score)) %>% 
  pull(Score) %>% 
  mean()

season_avg_accuracy <- df %>% 
  summarise(ShotsAccuracy = mean(ShotsAccuracy, na.rm = TRUE)) %>% 
  pull(ShotsAccuracy)

# Filter teams with single manager
teams <- df %>% 
  group_by(Team) %>% 
  summarise(L = length(unique(Manager))) %>% 
  ungroup() %>% 
  filter(L == 1) %>% 
  pull(Team)

# Create plot
df %>% 
  filter(Team %in% teams) %>% 
  group_by(Manager, Team) %>% 
  summarise(
    Score = sum(Score),
    ShotsAccuracy = mean(ShotsAccuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = Score, y = ShotsAccuracy, label = paste0(Manager, " (", Team, ")"))) +
  # Quadrant backgrounds
  annotate("rect", xmin = 20, xmax = season_avg_score, 
           ymin = 0.275, ymax = season_avg_accuracy, 
           fill = "#FFCDD2", alpha = 0.3) +
  annotate("rect", xmin = season_avg_score, xmax = 110, 
           ymin = season_avg_accuracy, ymax = 0.4, 
           fill = "#C8E6C9", alpha = 0.4) +
  annotate("rect", xmin = 20, xmax = season_avg_score, 
           ymin = season_avg_accuracy, ymax = 0.4, 
           fill = "#FFF9C4", alpha = 0.3) +
  annotate("rect", xmin = season_avg_score, xmax = 110, 
           ymin = 0.275, ymax = season_avg_accuracy, 
           fill = "#FFF9C4", alpha = 0.3) +
  # Reference lines
  geom_hline(yintercept = season_avg_accuracy, linetype = "dashed", 
             color = "gray40", linewidth = 0.8) +
  geom_vline(xintercept = season_avg_score, linetype = "dashed", 
             color = "gray40", linewidth = 0.8) +
  # Season average point
  geom_point(aes(x = season_avg_score, y = season_avg_accuracy), 
             size = 5, color = "#F57C00", shape = 21, fill = "white", stroke = 2) +
  annotate("text", x = season_avg_score + 4, y = season_avg_accuracy + 0.006, 
           label = "Season Average", color = "gray20", size = 4, 
           hjust = 0, fontface = "bold", family = "roboto") +
  # Data points
  geom_point(size = 4, alpha = 0.7, color = "#2D5F4F") +
  geom_text_repel(
    size = 3.5,
    family = "roboto",
    box.padding = 0.5,
    point.padding = 0.3,
    segment.color = "gray50",
    segment.size = 0.3,
    max.overlaps = Inf,
    min.segment.length = 0
  ) +
  scale_x_continuous(limits = c(20, 110), 
                     breaks = seq(20, 110, 15),
                     expand = expansion(mult = 0.05)) +
  scale_y_continuous(limits = c(0.275, 0.4),
                     labels = scales::percent_format(accuracy = 0.1),
                     breaks = seq(0.28, 0.40, 0.02),
                     expand = expansion(mult = 0.02)) +
  labs(
    title = "Manager Performance: Goals vs Shot Accuracy",
    subtitle = "Teams with single manager only | Green quadrant = above average on both metrics",
    x = "Total Goals Scored",
    y = "Average Shot Accuracy"
  ) +
  theme(base_family = "roboto") +
  theme(
    text = element_text(family = "roboto"),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "roboto"),
    plot.subtitle = element_text(size = 10, hjust = 0.5, family = "roboto", color = "gray40"),
    axis.title = element_text(size = 12, face = "bold", family = "roboto"),
    axis.text = element_text(size = 11, family = "roboto", color = "gray30"),
    axis.line = element_line(color = "gray40", linewidth = 0.5),
    axis.ticks = element_line(color = "gray40", linewidth = 0.3),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray50", fill = NA, linewidth = 0.5),
    plot.margin = margin(15, 15, 15, 15)
  )


```

# Conclusions

Data analysis is so very enjoyable and only more so when it reminds me of Liverpool's stunning 2019/20 season. YNWA.
